Iterative Monomorphisation

### 2024-01-31 iterative_monomorphisation.tex
applied some of the changes discussed with Jasmin

\author[1,2]{Tanguy Bozec}[%
email=tanguy.bozec@ens-paris-saclay.fr,
]
\author[2]{Jasmin Blanchette}[%
email=jasmin.blanchette@ifi.lmu.de,
%url=https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette_de.html,
]
\address[1]{Departement d'informatique, École normale supérieure Paris-Saclay, France} % TODO dpt info
\address[2]{Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}

@@

This paper will first give some general background and describe the logical framework. Then a high-level description of the algorithm will be presented. A more detailed low-level description of the algorithm corresponding to its implementation will follow.

Finally, the evaluation of the implementation in Zipperposition will be discussed.

\subsection{Context and motivation}
Automatic theorem provers have many applications, and are particularly useful as backends to proof assistants. Speed is one of the main limitations of the use of theorem provers in this setting. Indeed, a prover can often be expected to run for no more than a few seconds for a given instance of a problem.

@@

%basic idea of the algorithm
\subsection{Algorithm overview}
A given problem is seen as a set of polymorphic clauses. Type variables of the problem are quantified at the clause level. The type arguments of the problem are collected and the polymorphic type arguments are matched against the monomorphic ones. This yields new type arguments, both polymorphic and monomorphic, the process is then repeated making use of the newly derived type arguments.

@@

\subsection{Evaluation results}

@@

\begin{table}[th]
\caption{Monomorphic type argument generation}
\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
   \toprule
   & \multicolumn{3}{c}{floor 1} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 9}\\
   \cmidrule(l){2-10}
   \multirow{2}{2em}{cap} & \multicolumn{9}{c}{multiplier} \\
      & 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0\\
    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
    25   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    50   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    100  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    none & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \bottomrule
\end{tabular}
\end{table}

@@

If the floor and multiplier are set to 0 and 0.0, no new polymorphic type arguments will be generated, as a result the cap will be irrelevant.

\begin{table}[th]
\caption{Polymorphic type argument generation}
\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
   \toprule
   & \multicolumn{3}{c}{floor 0} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 6}\\
   \cmidrule(l){2-10}
   \multirow{2}{2em}{cap} & \multicolumn{9}{c}{multiplier} \\
      & 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0 \\
    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
    10   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    20   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    50   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    none & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[th]
\caption{Clause generation}
\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
   \toprule
   & \multicolumn{3}{c}{cap 500} & \multicolumn{3}{c}{cap 2000} & \multicolumn{3}{c}{no cap}\\
   \cmidrule(l){2-10}
   \multirow{2}{4em}{E timeout} & \multicolumn{9}{c}{multiplier} \\
      & 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0 \\
    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
    2s   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    5s   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    10s  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    none & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \bottomrule
\end{tabular}
\end{table}
The next table gathers the bounds linked to substitutions generation, the first one limits the number of monomorphising substitutions that are generated per clause during the final phase. The second one gives an absolute cap on the overall number of substitutions generated at each clause during the type argument generation.

\begin{table}[th]
\caption{Substitution generation}
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
   \toprule
   \multirow{2}{4em}{subst cap} & \multicolumn{4}{c}{mono subst}\\
   & 2 & 5 & 7 & 10\\
   \midrule
   50   & 0 & 0 & 0 & 0\\
   100  & 0 & 0 & 0 & 0\\
   500  & 0 & 0 & 0 & 0\\
   none & 0 & 0 & 0 & 0\\

@@

\begin{table}[th]
% not super happy with caption
\caption{Problem depth}
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
   \toprule
   \multirow{2}{3em}{loop nb} & \multicolumn{4}{c}{E call step}\\
   & 0 & 15 & 45 & 90\\
   \midrule
   2 & 0 & 0 & 0 & 0\\
   3 & 0 & 0 & 0 & 0\\
   4 & 0 & 0 & 0 & 0\\
   5 & 0 & 0 & 0 & 0\\

% We have two options, either use the current tables and make a tiered old/new test structure, or test everyting at once, which would be possible (32 options vs 16 with the tiered approach) however the results would most likely not be very pleasant to present, a highlight would have to be selected.

% final tests to make sure everything plays well together
\begin{table}[ht]
\caption{Problem size refinement}
\centering\begin{tabular}{@{}lcccc@{}}
   \toprule
   & \multicolumn{4}{c}{substitution generation} \\
   & \multicolumn{2}{c}{old} & \multicolumn{2}{c}{new} \\
   \cmidrule(lr){2-5} 
   \multirow{2}{4.5em}{clause gen} & \multicolumn{4}{c}{problem depth} \\
   & old & new & old & new \\
   \cmidrule(lr){2-3} \cmidrule(lr){4-5}
   old & 0  & 0 & 0 & 0 \\
   new & 0  & 0 & 0 & 0\\
   \bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Type argument generation refinement}
\centering\begin{tabular}{@{}llcc@{}}
   \toprule
   && \multicolumn{2}{c}{mono bounds} \\
   && old & new\\
   \cmidrule(r){3-4}
   \multirow{2}{5.4em}{poly bounds}
   &\multicolumn{1}{l}{old} & 0  & 0 \\
   &\multicolumn{1}{l}{new} & 0  & 0 \\
   \bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Final bounds refinement}
\centering\begin{tabular}{@{}llcc@{}}
   \toprule
   && \multicolumn{2}{c}{type argument bounds} \\
   && old & new\\
   \cmidrule(r){3-4}
   \multirow{2}{5.4em}{problem size bounds}
   &\multicolumn{1}{l}{old} & 0  & 0 \\
   &\multicolumn{1}{l}{new} & 0  & 0 \\
   \bottomrule
\end{tabular}
\end{table}

>>>>

\author[1,2]{Tanguy Bozec}[%
email=tanguy.bozec@ens-paris-saclay.fr,
]
\author[2]{Jasmin Blanchette}[%
email=jasmin.blanchette@ifi.lmu.de,
%url=https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette_de.html,
]
\address[1]{ENS Paris-Saclay, Université Paris-Saclay, France}
\address[2]{Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}

@@

Automatic theorem provers have many applications, and are particularly useful as backends to proof assistants. Speed is one of the main limitations of the use of theorem provers in this setting. Indeed, a prover can often be expected to run for no more than a few seconds for a given instance of a problem.

@@

%basic idea of the algorithm
A given problem is seen as a set of polymorphic clauses. Type variables of the problem are quantified at the clause level. The type arguments of the problem are collected and the polymorphic type arguments are matched against the monomorphic ones. This yields new type arguments, both polymorphic and monomorphic, the process is then repeated making use of the newly derived type arguments.

@@

@@

\begin{table}[th]
\caption{Evaluation of bounds for monomorphic type argument generation}
\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
   \toprule
   & &&& \multicolumn{3}{c}{floor} \\
   & \multicolumn{3}{c}{1} & \multicolumn{3}{c}{3} & \multicolumn{3}{c}{9}\\
   \cmidrule(l){2-10}
   & &&& \multicolumn{3}{c}{multiplier} \\
   \multirow{1}{2em}{cap} & 0 & 1 & 2 & 0 & 1 & 2 & 0 & 1 & 2\\
    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
    25   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    50   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    100  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \(\infty\) & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \bottomrule
\end{tabular}
\end{table}

@@

If the floor and multiplier are set to 0 and 0, no new polymorphic type arguments will be generated, as a result the cap will be irrelevant.

\begin{table}[th]
\caption{Evaluation of bounds for polymorphic type argument generation}
\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
   \toprule
   & &&& \multicolumn{3}{c}{floor} \\
   & \multicolumn{3}{c}{0} & \multicolumn{3}{c}{3} & \multicolumn{3}{c}{6}\\
   \cmidrule(l){2-10}
   & &&& \multicolumn{3}{c}{multiplier} \\
    \multirow{1}{2em}{cap} & 0 & 0.5 & 1 & 0 & 0.5 & 1 & 0 & 0.5 & 1 \\
    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
    10   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    20   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    40   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \(\infty\) & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \bottomrule
\end{tabular}
\end{table}

\begin{table}[th]
\caption{Evaluation of bounds for substitution generation}
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
   \toprule
   & \multicolumn{4}{c}{mono subst}\\
   \multirow{1}{4em}{subst cap} & 2 & 5 & 7 & 10\\
   \midrule
   50   & 0 & 0 & 0 & 0\\
   100  & 0 & 0 & 0 & 0\\
   500  & 0 & 0 & 0 & 0\\
   \(\infty\) & 0 & 0 & 0 & 0\\

@@

\begin{table}[th]
\caption{Evaluation of bounds directly related to the size of the output problem}
\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
   \toprule
   & &&& \multicolumn{3}{c}{cap} \\
   & \multicolumn{3}{c}{500} & \multicolumn{3}{c}{2000} & \multicolumn{3}{c}{\(\infty\)}\\
   \cmidrule(l){2-10}
   & &&& \multicolumn{3}{c}{multiplier} \\
    \multirow{1}{5.4em}{E timeout (s)} & 1 & 2 & 3 & 1 & 2 & 3 & 1 & 2 & 3 \\
    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
    2   &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    5   &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    10  &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \(\infty\)& 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \bottomrule
\end{tabular}
\end{table}
The next table gathers the bounds linked to substitutions generation, the first one limits the number of monomorphising substitutions that are generated per clause during the final phase. The second one gives an absolute cap on the overall number of substitutions generated at each clause during the type argument generation.

\begin{table}[th]
% not super happy with caption
\caption{Evaluation of parameters indirectly related to the size of the output problem}
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
   \toprule
   & \multicolumn{4}{c}{loop nb} \\
   \multirow{1}{4.5em}{E call step} & 2 & 3 & 4 & 5\\
   \midrule
   0 & 0 & 0 & 0 & 0\\
   15 & 0 & 0 & 0 & 0\\
   45 & 0 & 0 & 0 & 0\\
   90 & 0 & 0 & 0 & 0\\

@@
